Using functor categories to generate intermediate code with Agda

### 2025-05-17 src/compiler.lagda.tex
draft-ver-2

\usepackage{agda}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{mathabx}

\usepackage{unicode-math}
\setmathfont{XITS Math}

\usepackage{newunicodechar}
\newunicodechar{≤}{\ensuremath{\leq}}
\newunicodechar{≡}{\ensuremath{\equiv}}
\newunicodechar{∸}{\ensuremath{\dotdiv}}
\newunicodechar{∀}{\ensuremath{\forall}}
\newunicodechar{→}{\ensuremath{\rightarrow}}
\newunicodechar{ℕ}{\ensuremath{\mathbb{N}}}
\newunicodechar{ℤ}{\ensuremath{\mathbb{Z}}}
\newunicodechar{δ}{\ensuremath{\delta}}
\newunicodechar{∈}{\ensuremath{\in}}
\newunicodechar{⊢}{\ensuremath{\vdash}}
\newunicodechar{⇒}{\ensuremath{\Rightarrow}}
\newunicodechar{⟶}{\ensuremath{\longrightarrow}}
\newunicodechar{ₛ}{\ensuremath{\textsubscript{s}}}
\newunicodechar{ᵛ}{\ensuremath{\textsuperscript{v}}}

@@

_⇒ₛ_ : (SD → Set) → (SD → Set) → SD → Set
(P ⇒ₛ Q) sd = ∀{sd'} → (sd ≤ₛ sd') → P sd' → Q sd'

@@

get-var : ∀ {Γ A sd} → A ∈ Γ → ⟦ Γ ⟧ctx sd → ⟦ A ⟧ty sd
get-var Zero     (_ , a) = a
get-var (Suc x) (γ , _) = get-var x γ

-- Functorality
fmap-⇒ : ∀ {P Q sd sd'} → (P ⇒ₛ Q) sd → sd ≤ₛ sd' → (P ⇒ₛ Q) sd'
fmap-⇒ θ p p' x = θ (≤ₛ-trans p p') x

fmap-Compl : ∀ {sd sd'} → Compl sd → sd ≤ₛ sd' → Compl sd'
fmap-Compl {sd} c (<-f f<f') = popto sd (<-f f<f') c
fmap-Compl {⟨ f , d ⟩} {⟨ f , d' ⟩} c (≤-d d≤d') = 
    adjustdisp-dec ((d' - d) d≤d') (-→≤ d≤d') 
        (sub I (-ₛ≡ 
            {n = (d' - d) d≤d'} 
                (n-[n-m]≡m d≤d')) c)

fmap-L : ∀ {sd sd'} → L sd → sd ≤ₛ sd' → L sd'
fmap-L (l-var sdᵛ sdᵛ≤ₛsd) sd≤ₛsd' = l-var sdᵛ (≤ₛ-trans sdᵛ≤ₛsd sd≤ₛsd')
fmap-L (l-sbrs) _ = l-sbrs

fmap-S : ∀ {sd sd'} → S sd → sd ≤ₛ sd' → S sd'
fmap-S (s-l l) sd≤ₛsd' = s-l (fmap-L l sd≤ₛsd')
fmap-S (s-lit lit) _ = s-lit lit

fmap-A : ∀ {A sd sd'} → ⟦ A ⟧ty sd → sd ≤ₛ sd' → ⟦ A ⟧ty sd'
fmap-A {comm}  = fmap-⇒ {Compl} {Compl}
fmap-A {intexp} = fmap-⇒ {Intcompl} {Compl}
fmap-A {intacc} = fmap-⇒ {Compl} {Intcompl}
fmap-A {intvar} ( e , a ) sd≤ₛsd' = ( fmap-A {intexp} e sd≤ₛsd' , fmap-A {intacc} a sd≤ₛsd' )
fmap-A {A ⇒ B} = fmap-⇒ {⟦ A ⟧ty} {⟦ B ⟧ty}

fmap-Γ : ∀ {Γ sd sd'} → ⟦ Γ ⟧ctx sd → sd ≤ₛ sd' → ⟦ Γ ⟧ctx sd'
fmap-Γ {·} unit _ = unit
fmap-Γ {Γ , A} (γ , a) p = fmap-Γ γ p , fmap-A {A} a p

sd≤ₛsd'→sd≤ₛsd'-ₛ[d'-[suc-d]] : ∀ {sd sd'} → sd ≤ₛ sd' → (δ₁≤δ₂ : suc (SD.d sd) ≤ SD.d sd') → sd ≤ₛ ((sd' -ₛ ((SD.d sd' - (suc (SD.d sd))) δ₁≤δ₂)) (-→≤ δ₁≤δ₂))
sd≤ₛsd'→sd≤ₛsd'-ₛ[d'-[suc-d]] {⟨ f , _ ⟩} {⟨ f' , _ ⟩} (<-f f<f') _ = <-f f<f'
sd≤ₛsd'→sd≤ₛsd'-ₛ[d'-[suc-d]] {⟨ f , d ⟩} {⟨ f , d' ⟩} (≤-d d≤d') δ₁≤δ₂ = ≤-d (suc-d≤d'→d≤d'-[d'-[suc-d]] δ₁≤δ₂)

assign : (sd : SD) → (sd' : SD) → (S ⇒ₛ Compl) sd → sd ≤ₛ sd' → R sd' → I sd'
assign ⟨ f , d ⟩ ⟨ f' , d' ⟩ β sd≤ₛsd' r with (≤-compare {suc d} {d'})
... | leq δ₁≤δ₂ = assign-dec ((d' - (suc d)) δ₁≤δ₂) 
                    (-→≤ δ₁≤δ₂) (l-var ⟨ f , d ⟩ (sd≤ₛsd'→sd≤ₛsd'-ₛ[d'-[suc-d]] sd≤ₛsd' δ₁≤δ₂)) r 
                    (β ((sd≤ₛsd'→sd≤ₛsd'-ₛ[d'-[suc-d]] sd≤ₛsd' δ₁≤δ₂)) 
                        (s-l (l-var ⟨ f , d ⟩ ((sd≤ₛsd'→sd≤ₛsd'-ₛ[d'-[suc-d]] sd≤ₛsd' δ₁≤δ₂)))))

... | geq δ₂≤δ₁ = assign-inc (((suc d) - d') δ₂≤δ₁) 
                    (l-var ⟨ f , d ⟩ (≤ₛ-trans sd≤ₛsd' +ₛ→≤ₛ)) r 
                    (β ((≤ₛ-trans sd≤ₛsd' +ₛ→≤ₛ)) 
                        (s-l (l-var ⟨ f , d ⟩ ((≤ₛ-trans sd≤ₛsd' +ₛ→≤ₛ)))))

use-temp : ∀ {sd sd'} → (S ⇒ₛ Compl) sd → sd ≤ₛ sd' → R sd' → I sd'
use-temp β sd≤ₛsd' (r-s s) = β sd≤ₛsd' s
use-temp {sd} {sd'} β sd≤ₛsd' (r-unary uop s) = 
    assign sd sd' β sd≤ₛsd' (r-unary uop s)
use-temp {sd} {sd'} β sd≤ₛsd' (r-binary s₁ bop s₂) = 
    assign sd sd' β sd≤ₛsd' (r-binary s₁ bop s₂)

⟦_⟧ : ∀ {Γ A} → (e : Γ ⊢ A) → (sd : SD) → ⟦ Γ ⟧ctx sd → ⟦ A ⟧ty sd
⟦ Var x ⟧ _ γ = get-var x γ
⟦ Lambda f ⟧ sd γ {sd' = sd'} p a = ⟦ f ⟧ sd' (fmap-Γ γ p , a) 
⟦ App f e ⟧ sd γ = ⟦ f ⟧ sd γ (≤-d ≤-refl) (⟦ e ⟧ sd γ)
⟦ Skip ⟧ _ _ _ γ = γ
⟦ Seq c₁ c₂ ⟧ sd γ sd' p = ⟦ c₁ ⟧ sd γ sd' (⟦ c₂ ⟧ sd γ sd' p)
⟦ Lit i ⟧ _ _ _ κ = κ ≤ₛ-refl (r-s (s-lit i))
⟦ Neg e ⟧ sd γ p κ = ⟦ e ⟧ sd γ p (use-temp λ p s → κ p (r-unary UNeg s))    
⟦ Plus e₁ e₂ ⟧ sd γ p κ = ⟦ e₁ ⟧ sd γ p (use-temp (λ p' s₁ → ⟦ e₂ ⟧ sd γ (≤ₛ-trans p p') (use-temp (λ p'' s₂ →  κ (≤ₛ-trans p' p'') (r-binary (fmap-S s₁ p'') BPlus s₂)))))
\end{code}
\end{prev}
\end{document}

>>>>

\usepackage{agda}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{mathabx}
\usepackage{fontspec}
\setmainfont{Linux Libertine O}
\setsansfont{Linux Biolinum O}

\usepackage{newunicodechar}
\newunicodechar{ℕ}{\ensuremath{\mathbb{N}}}
\newunicodechar{ℤ}{\ensuremath{\mathbb{Z}}}
\newunicodechar{λ}{\ensuremath{\lambda}}
\newunicodechar{∸}{\ensuremath{\dotdiv}}
\newunicodechar{→}{\ensuremath{\rightarrow}}
\newunicodechar{⇒}{\ensuremath{\Rightarrow}}
\newunicodechar{⟶}{\ensuremath{\longrightarrow}}
\newunicodechar{ᵛ}{\ensuremath{\textsuperscript{v}}}
\newunicodechar{ʳ}{\ensuremath{\textsuperscript{r}}}
\newunicodechar{⟨}{\ensuremath{\langle}}
\newunicodechar{⟩}{\ensuremath{\rangle}}

@@

_⇒ₛ_ : (SD → Set) → (SD → Set) → SD → Set
(P ⇒ₛ Q) sd = ∀{sd′} → (sd ≤ₛ sd′) → P sd′ → Q sd′

@@

⟦_⟧var : ∀ {Γ A sd} → A ∈ Γ → ⟦ Γ ⟧ctx sd → ⟦ A ⟧ty sd
⟦ Zero ⟧var (_ , a) = a
⟦ Suc b ⟧var (γ , _) =  ⟦ b ⟧var γ

⟦_⟧sub : ∀ {A A′ sd} → A ≤: A′ → ⟦ A ⟧ty sd → ⟦ A′ ⟧ty sd
⟦ ≤:-refl ⟧sub a = a
⟦ ≤:-trans A≤:A′ A′≤:A″ ⟧sub a = ⟦ A′≤:A″ ⟧sub (⟦ A≤:A′ ⟧sub a)
⟦ ≤:-fn A≤:A′ B′≤:B ⟧sub a = 
    λ sd≤ₛsd′ a′ → ⟦ B′≤:B ⟧sub (a sd≤ₛsd′ (⟦ A≤:A′ ⟧sub a′))
⟦ var-≤:-exp ⟧sub (exp , acc) = exp
⟦ var-≤:-acc ⟧sub (exp , acc) = acc

-- Functorial mapping

fmap-I : ∀ {sd sd′} → I sd → sd ≤ₛ sd′ → I sd′
fmap-I {sd} c (<-f f<f′) = popto sd (<-f f<f′) c
fmap-I {⟨ f , d ⟩} {⟨ f , d′ ⟩} c (≤-d d≤d′) = 
    adjustdisp-dec ((d′ – d) d≤d′) (–→≤ d≤d′) 
        (I-sub {n = (d′ – d) d≤d′} (n–[n–m]≡m d≤d′) c)

fmap-L : ∀ {sd sd′} → L sd → sd ≤ₛ sd′ → L sd′
fmap-L (l-var sdᵛ sdᵛ≤ₛsd) sd≤ₛsd′ = l-var sdᵛ (≤ₛ-trans sdᵛ≤ₛsd sd≤ₛsd′)
fmap-L (l-sbrs) _ = l-sbrs

fmap-S : ∀ {sd sd′} → S sd → sd ≤ₛ sd′ → S sd′
fmap-S (s-l l) sd≤ₛsd′ = s-l (fmap-L l sd≤ₛsd′)
fmap-S (s-lit lit) _ = s-lit lit

fmap-⇒ : ∀ {P Q sd sd′} → (P ⇒ₛ Q) sd → sd ≤ₛ sd′ → (P ⇒ₛ Q) sd′
fmap-⇒ P⇒Q sd≤ₛsd′ sd′≤ₛsd″ p = P⇒Q (≤ₛ-trans sd≤ₛsd′ sd′≤ₛsd″) p

fmap-ty : ∀ {A sd sd′} → ⟦ A ⟧ty sd → sd ≤ₛ sd′ → ⟦ A ⟧ty sd′
fmap-ty {comm}  = fmap-⇒ {Compl} {Compl}
fmap-ty {intexp} = fmap-⇒ {Intcompl} {Compl}
fmap-ty {intacc} = fmap-⇒ {Compl} {Intcompl}
fmap-ty {intvar} ( exp , acc ) sd≤ₛsd′ = 
    ( fmap-ty {intexp} exp sd≤ₛsd′ , fmap-ty {intacc} acc sd≤ₛsd′ )
fmap-ty {A ⇒ B} = fmap-⇒ {⟦ A ⟧ty} {⟦ B ⟧ty}

fmap-ctx : ∀ {Γ sd sd′} → ⟦ Γ ⟧ctx sd → sd ≤ₛ sd′ → ⟦ Γ ⟧ctx sd′
fmap-ctx {·} unit _ = unit
fmap-ctx {Γ , A} (γ , a) p = fmap-ctx γ p , fmap-ty {A} a p

sd≤ₛsd′→sd≤ₛsd′–ₛ[d′–[suc-d]] : ∀ {sd sd′} → sd ≤ₛ sd′ 
    → (δ₁≤δ₂ : suc (SD.d sd) ≤ SD.d sd′) 
    → sd ≤ₛ ((sd′ –ₛ ((SD.d sd′ – (suc (SD.d sd))) δ₁≤δ₂)) (–→≤ δ₁≤δ₂))
sd≤ₛsd′→sd≤ₛsd′–ₛ[d′–[suc-d]] {⟨ f , _ ⟩} {⟨ f′ , _ ⟩} (<-f f<f′) _ 
    = <-f f<f′
sd≤ₛsd′→sd≤ₛsd′–ₛ[d′–[suc-d]] {⟨ f , d ⟩} {⟨ f , d′ ⟩} (≤-d d≤d′) δ₁≤δ₂ 
    = ≤-d (suc-d≤d′→d≤d′–[d′–[suc-d]] δ₁≤δ₂)

new-intvar : ∀ sd → ⟦ intvar ⟧ty sd
new-intvar sd = ( exp , acc )
    where
        exp : ⟦ intexp ⟧ty sd
        exp sd≤ₛsd′ β = β ≤ₛ-refl (r-s (s-l (l-var sd sd≤ₛsd′)))
        acc : ⟦ intacc ⟧ty sd
        acc {sd′ = sd′} sd≤ₛsd′ κ (≤-d {d = d′} {d′ = d″} d′≤d″) r
            = assign-dec 
                ((d″ – d′) d′≤d″) (–→≤ d′≤d″) 
                (l-var sd
                   (sub-sd≤ₛ (–ₛ≡ {n≤d′ = –→≤ d′≤d″} (n–[n–m]≡m d′≤d″)) sd≤ₛsd′))
                r
                (I-sub {n = (d″ – d′) d′≤d″}(n–[n–m]≡m d′≤d″) κ)
        acc {sd′ = sd′} sd≤ₛsd′ κ (<-f f<f′) r 
            = assign-inc 0 (l-var _ ≤ₛ-refl) r (fmap-I κ (<-f f<f′))

assign : (sd : SD) → (sd′ : SD) → (S ⇒ₛ Compl) sd 
            → sd ≤ₛ sd′ → R sd′ → I sd′
assign ⟨ f , d ⟩ ⟨ f′ , d′ ⟩ β sd≤ₛsd′ r with (≤-compare {suc d} {d′})
... | leq δ₁≤δ₂ 
        = assign-dec 
            ((d′ – (suc d)) δ₁≤δ₂) (–→≤ δ₁≤δ₂) 
              (l-var ⟨ f , d ⟩ 
                (sd≤ₛsd′→sd≤ₛsd′–ₛ[d′–[suc-d]] sd≤ₛsd′ δ₁≤δ₂)) 
              r 
              (β ((sd≤ₛsd′→sd≤ₛsd′–ₛ[d′–[suc-d]] sd≤ₛsd′ δ₁≤δ₂)) 
                (s-l (l-var ⟨ f , d ⟩ 
                  ((sd≤ₛsd′→sd≤ₛsd′–ₛ[d′–[suc-d]] sd≤ₛsd′ δ₁≤δ₂)))))

... | geq δ₂≤δ₁ = assign-inc (((suc d) – d′) δ₂≤δ₁) 
                  (l-var ⟨ f , d ⟩ (≤ₛ-trans sd≤ₛsd′ +ₛ→≤ₛ)) r 
                  (β ((≤ₛ-trans sd≤ₛsd′ +ₛ→≤ₛ)) 
                    (s-l (l-var ⟨ f , d ⟩ ((≤ₛ-trans sd≤ₛsd′ +ₛ→≤ₛ)))))

use-temp : ∀ {sd sd′} → (S ⇒ₛ Compl) sd → sd ≤ₛ sd′ → R sd′ → I sd′
use-temp β sd≤ₛsd′ (r-s s) = β sd≤ₛsd′ s
use-temp {sd} {sd′} β sd≤ₛsd′ (r-unary uop s) = 
    assign sd sd′ β sd≤ₛsd′ (r-unary uop s)
use-temp {sd} {sd′} β sd≤ₛsd′ (r-binary s₁ bop s₂) = 
    assign sd sd′ β sd≤ₛsd′ (r-binary s₁ bop s₂)

⟦_⟧ : ∀ {Γ A} → Γ ⊢ A → (sd : SD) → ⟦ Γ ⟧ctx sd → ⟦ A ⟧ty sd
⟦ Var a ⟧ sd γ = ⟦ a ⟧var γ

⟦ Sub a A≤:B ⟧ sd γ = ⟦ A≤:B ⟧sub (⟦ a ⟧ sd γ)

⟦ Lambda f ⟧ sd γ {sd′ = sd′} sd≤ₛsd′ a = ⟦ f ⟧ sd′ (fmap-ctx γ sd≤ₛsd′ , a)

⟦ App f e ⟧ sd γ = ⟦ f ⟧ sd γ (≤-d ≤-refl) (⟦ e ⟧ sd γ)

⟦ Skip ⟧ sd γ sd≤ₛsd′ κ = κ

⟦ Seq c₁ c₂ ⟧ sd γ sd≤ₛsd′ κ = ⟦ c₁ ⟧ sd γ sd≤ₛsd′ (⟦ c₂ ⟧ sd γ sd≤ₛsd′ κ)

⟦ NewVar c ⟧ sd γ {sd′ = sd′} sd≤ₛsd′ κ = 
    assign-inc 1 
        (l-var sd′ (≤-d +→≤))
        (r-s (s-lit (pos 0)))
        (⟦ c ⟧
            (sd′ +ₛ 1)
            (fmap-ctx {Γ = _ , intvar} 
                ((fmap-ctx γ sd≤ₛsd′ , new-intvar sd′)) 
                (+ₛ→≤ₛ {sd′} {1}))
            ≤ₛ-refl
            (adjustdisp-dec 1 +→≤ʳ
                (I-sub {d′ = SD.d sd′ + 1} {n = 1} 
                    (n+m–m≡n {m = 1}) κ)))

⟦ Assign a e ⟧ sd γ sd≤ₛsd′ κ = ⟦ e ⟧ sd γ sd≤ₛsd′ (⟦ a ⟧ sd γ sd≤ₛsd′ κ)

⟦ Lit i ⟧ sd γ sd≤ₛsd′ β = β ≤ₛ-refl (r-s (s-lit i))

⟦ Neg e ⟧ sd γ sd≤ₛsd′ β = 
    ⟦ e ⟧ sd γ sd≤ₛsd′ (use-temp λ sd≤ₛsd′ s → β sd≤ₛsd′ (r-unary UNeg s))

⟦ Plus e₁ e₂ ⟧ sd γ sd≤ₛsd′ β = 
    ⟦ e₁ ⟧ sd γ sd≤ₛsd′ 
        (use-temp (λ sd′≤ₛsd″ s₁ → ⟦ e₂ ⟧ sd γ (≤ₛ-trans sd≤ₛsd′ sd′≤ₛsd″)
            (use-temp (λ sd″≤ₛsd‴ s₂ → β (≤ₛ-trans sd′≤ₛsd″ sd″≤ₛsd‴)
                (r-binary (fmap-S s₁ sd″≤ₛsd‴) BPlus s₂)))))

compile-closed : · ⊢ comm → I ⟨ 0 , 0 ⟩
compile-closed t = ⟦ t ⟧ ⟨ 0 , 0 ⟩ unit ≤ₛ-refl stop
\end{code}
\end{prev}
\end{document}
